---
title: "03 - augment"
format: html
editor: visual
---

## Load Libraries

```{r}
library("tidyverse")
library("dplyr")
```

## Load cleaned tables

```{r}
gen_exp_clean <- readRDS("../data/gen_exp_clean.rds")
phenotype_clean <- readRDS("../data/phenotype_clean.rds")
```

## Reshape gene expression table

The current gene expression dataset is in a wide format:\
rows = genes columns = samples\
For further analyses, it is more convenient to work in a long format: one row per (gene × sample) measurement. The dataset is now tidy.

```{r}
gen_exp_long <- gen_exp_clean |>
  pivot_longer(cols = -c(Ensembl_ID, sum_gene_exp),
               names_to = "sample",
               values_to = "expression"
               )

gen_exp_long |> 
  slice_sample(n = 10)
```

## Join clinical metadata

Now we merge the gene expression table with the phenotype table using the common identifier sample.

```{r}
# Merge expression data with phenotype metadata
full_data <- gen_exp_long |>
  left_join(phenotype_clean, by = "sample") |>
  relocate(sample, .before = Ensembl_ID)

full_data |> 
  slice_sample(n = 10)
```

We now have: biological measurements (gene-exp) + patient characteristics (phenotype).

## Create additional variables

We can also create new useful variables like:

-   A log-transformed expression variable (log2_expression), commonly used in transcriptomics to reduce skewness and easier interpretation.

-   A categorical age variable (age_cat), based on common thresholds in clinical studies.

```{r}
full_data <- full_data |>
  mutate(log2_expression = log2(expression + 1),
         age_cat = case_when(
           age_at_index.demographic < 50 ~ "young",
           age_at_index.demographic >= 50 & age_at_index.demographic < 65 ~ "middle",
           age_at_index.demographic >= 65 ~ "senior",
           TRUE ~ NA_character_)
         )

full_data |> 
  slice_sample(n = 10)
```

The +1 in the log transformation prevents issues when expression = 0. (goes to -inf with log)

NA_character\_ ensures that missing age values remain coded as valid character-type missing values in the age category variable.

## Save augmented dataset

```{r}
saveRDS(full_data, "../data/data_augmented.rds")
```
