---
title: "05 - analysis 1"
format: html
editor: visual
---

## Loading libraries and data

```{r}
library("tidyverse")
library("dplyr")
```

```{r}
# Load 3 tables
phenotype <- readRDS("../data/phenotype_clean.rds")
gen_exp <- readRDS("../data/gen_exp_clean.rds")
data_augmented <- readRDS("../data/data_augmented.rds")
```

```{r}
View(phenotype)
```

## Analysis of the disitrubtion of variables

age : dans describe

figo stage : We can see that most of the patients are in stage IIIC

```{r}
data_augmented |>
  group_by(sample) |>
  slice(1) |>
  filter(!is.na(figo_stage.diagnoses)) |>
  ggplot(aes(x = figo_stage.diagnoses)) +
    geom_bar() +
    labs(title = "distribution of FIGO stages",
       x = "FIGO stage",
       y = "number of patients")
```

alive vs dead : we can see that there is more dead patients than alive

```{r}
data_augmented |>
  group_by(sample) |>
  slice(1) |>
  ggplot(aes(x = alive)) +
    geom_bar() +
    labs(title = "distribution of alive patients",
       x = "alive",
       y = "number of patients")
```

primary vs recurent tumor : we can see that for most of the patients, the tumor is primary

```{r}
data_augmented |>
  group_by(sample) |>
  slice(1) |>
  ggplot(aes(x = tumor_descriptor.samples)) +
    geom_bar() +
    labs(title = "distribution of primary vs recurent tumor",
       x = "primary vs recurent",
       y = "number of patients")
```

types of treatment : the types of treatment are almost equally distributed

```{r}
data_augmented |>
  group_by(sample) |>
  slice(1) |>
  ggplot(aes(x = treatment_type.treatments.diagnoses)) +
    geom_bar() +
    scale_x_discrete(labels = c(
      "['Pharmaceutical Therapy, NOS', 'Radiation Therapy, NOS']" = "pharmaceutical before radiation",
      "['Radiation Therapy, NOS', 'Pharmaceutical Therapy, NOS']" = "radiation before pharmaceutical")) +
    labs(title = "distribution of ptypes of treatment",
       x = "types of treatment",
       y = "number of patients")
```

```{r}
gen_exp <- gen_exp |> 
  select(!sum_gene_exp)
```

## Biological relevance of most variant genes

In this section we will extract the top 10 genes by variance of FPKM, found their biological signification and plotted expression values across all samples.

### Finding most variant genes

```{r}
# Compute variance of expression per gene
gene_var <- gen_exp |>
  rowwise() |>
  mutate(var_expression = var(c_across(-Ensembl_ID), na.rm = TRUE), .before = 2) |>
  ungroup() |>
  arrange(desc(var_expression))

# Extract top 10 genes
genes_of_interest <- gene_var |> 
  slice_head(n = 10) |> 
  pull(Ensembl_ID) # Makes genes_of_interest a character vector with genes names


View(gene_var)
```

```{r}
genes_of_interest
```

The genes are listed using their Ensembl ID, which is a unique identifier assigned by the Ensembl projects. we will convert them to gene symbols to make them more understandable.

| Ensembl_ID        | Gene symbol |
|-------------------|-------------|
| ENSG00000211592.8 | IGKC        |
| ENSG00000211896.7 | IGHG1       |
| ENSG00000239951.1 | IGKV3-20    |
| ENSG00000211598.2 | IGKV4-1     |
| ENSG00000253755.1 | IGHGP       |
| ENSG00000211673.2 | IGLV3-1     |
| ENSG00000243466.1 | IGKV1-5     |
| ENSG00000157005.4 | SST         |
| ENSG00000211677.2 | IGLC2       |
| ENSG00000241351.3 | IGKV3-11    |

These are all immunoglobulin genes, except for SST gene. They are expressed by B cells and plasma cells, not by tumor cells. Their high variability likely reflects differences in immune infiltration across ovarian tumor samples.

This suggests that immune involvement varies substantially among patients, which is consistent with known heterogeneity in the tumor microenvironment. In other words, different patients' tumors contain different levels of immune cells, a result that is expected in ovarian cancer.

### Expression across samples

```{r}
# Boxplot of expression
gen_exp |>
  filter(Ensembl_ID %in% genes_of_interest) |>
  mutate(Ensembl_ID = factor(Ensembl_ID, levels = genes_of_interest)) |>
  pivot_longer(cols = -Ensembl_ID,
               names_to = "sample",
               values_to = "expression") |>
  ggplot(aes(x = Ensembl_ID, y = expression)) +
  geom_boxplot() +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(
    angle = 45,        
    hjust = 1,         
    size = 6           
    )
  ) +
  labs(
    title = "Expression of the 10 most variable genes across patients",
    y = "Expression"
  )
    
```

### Relationship with phenotype variables

```{r}
# Joining the genes of interest expressions with phenotype
joined <- gen_exp |>
  filter(Ensembl_ID %in% genes_of_interest) |>
  pivot_longer(cols = -Ensembl_ID,
               names_to = "sample",
               values_to = "expr") |>
  group_by(sample) |>
  summarize(immune_score = mean(expr)) |>
  left_join(phenotype, by = "sample")

```

After several attempts, we found no clear relationship between these genes expressions and any patients characteristics, such as age or alive status. These plots show the lack of association between the variables.

```{r}
# Immune genes vs age
ggplot(joined, aes(x = immune_score, y = age_at_index.demographic)) + 
  geom_point()

# Immune genese vs alive/dead
ggplot(joined, aes(x = immune_score, y = alive)) + 
  geom_boxplot()

# Immune genes vs age
ggplot(joined, aes(x = immune_score, y = race.demographic)) + 
  geom_boxplot()
```
