---
title: "05 - analysis 1"
format: html
editor: visual
---

### Loading libraries and data

```{r}
library("tidyverse")
```

```{r}
# Load 3 tables
phenotype <- readRDS("../data/phenotype_clean.rds")
gen_exp <- readRDS("../data/gen_exp_clean.rds")
data_augmented <- readRDS("../data/data_augmented.rds")
```

## Distribution of the variables

age : dans describe

figo stage : We can see that most of the patients are in stage IIIC

```{r}
data_augmented |>
  group_by(sample) |>
  slice(1) |>
  filter(!is.na(figo_stage.diagnoses)) |>
  ggplot(aes(x = figo_stage.diagnoses)) +
    geom_bar() +
    labs(title = "distribution of FIGO stages",
       x = "FIGO stage",
       y = "number of patients")
```

alive vs dead : we can see that there is more dead patients than alive

```{r}
data_augmented |>
  group_by(sample) |>
  slice(1) |>
  ggplot(aes(x = alive)) +
    geom_bar() +
    labs(title = "distribution of alive patients",
       x = "alive",
       y = "number of patients")
```

primary vs recurent tumor : we can see that for most of the patients, the tumor is primary

```{r}
data_augmented |>
  group_by(sample) |>
  slice(1) |>
  ggplot(aes(x = tumor_descriptor.samples)) +
    geom_bar() +
    labs(title = "distribution of primary vs recurent tumor",
       x = "primary vs recurent",
       y = "number of patients")
```

types of treatment : the types of treatment are almost equally distributed

```{r}
data_augmented |>
  group_by(sample) |>
  slice(1) |>
  ggplot(aes(x = treatment_type.treatments.diagnoses)) +
    geom_bar() +
    scale_x_discrete(labels = c(
      "['Pharmaceutical Therapy, NOS', 'Radiation Therapy, NOS']" = "pharmaceutical before radiation",
      "['Radiation Therapy, NOS', 'Pharmaceutical Therapy, NOS']" = "radiation before pharmaceutical")) +
    labs(title = "distribution of ptypes of treatment",
       x = "types of treatment",
       y = "number of patients")
```

```{r}
# Select most variable gene
gene_var <- full_data |>
  group_by(Ensembl_ID) |>
  summarise(var_log2 = var(log2_expression, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(var_log2))

# Choose the top variable gene
gene_x <- gene_var$Ensembl_ID[1]

# Filter data for this gene and compute survival duration
df_gene <- full_data |>
  filter(Ensembl_ID == gene_x) |>
  mutate(
    survival_days = case_when(
      alive == TRUE  ~ days_to_last_follow_up.diagnoses,
      alive == FALSE ~ days_to_death.demographic,
      TRUE ~ NA_real_
    )
  )

# -------------------------------------------------------------
# Boxplot: Expression vs Alive
# -------------------------------------------------------------
ggplot(df_gene, aes(x = alive, y = log2_expression)) +
  geom_boxplot(fill = "#a6cee3") +
  labs(
    title = paste("Expression of gene", gene_x, "in alive vs dead patients"),
    x = "Alive status",
    y = "log2(expression)"
  ) +
  theme_minimal()

# -------------------------------------------------------------
# Scatter plot + Regression: Survival vs Expression
# -------------------------------------------------------------
ggplot(df_gene, aes(x = log2_expression, y = survival_days, color = alive)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = paste("Survival time vs expression for gene", gene_x),
    x = "log2(expression)",
    y = "Survival (days)"
  ) +
  theme_minimal()

# -------------------------------------------------------------
# Linear model summary
# -------------------------------------------------------------
lm(survival_days ~ log2_expression, data = df_gene) |> summary()
```

