---
title: "09_analysis5"
format: html
editor: visual
---

## Aim

In this analysis, we explore whether transcriptomic profiles group patients into meaningful clusters. We perform hierarchical clustering based on the 100 most variable genes across samples and visualise the resulting structure using a heatmap with clinical annotations.

## Setup

```{r}
library("tidyverse")
```

## Load augmented data

```{r}
full_data <- readRDS("../data/data_augmented.rds")

full_data |>
  slice_sample(n = 10)

```

## Select the 100 most variable genes

We compute the variance of log2 expression for each gene and select the top 100 most variable ones.

```{r}
gene_var <- full_data |>
  group_by(Ensembl_ID) |>
  summarise(
    var_log2 = var(log2_expression, na.rm = TRUE),
    .groups  = "drop") |>
  arrange(desc(var_log2))

top100_genes <- gene_var |>
  slice_head(n = 100) |>
  pull(Ensembl_ID)

length(top100_genes)
```

## Create a sample x gene matrix

We reshape the dataset to obtain a wide expression matrix and attach the clinical annotation `alive` for colouring later.

```{r}
expr_wide <- full_data |>
  filter(Ensembl_ID %in% top100_genes) |>
  select(sample, Ensembl_ID, log2_expression) |>
  distinct() |>
  pivot_wider(names_from = Ensembl_ID,
              values_from = log2_expression)

pheno <- full_data |>
  select(sample, alive) |>
  distinct()

expr_wide <- expr_wide |>
  left_join(pheno, by = "sample")

expr_wide |>
  slice_head(n = 5)
```

## Prepare expression matrix for clustering

We extract only numeric gene columns and convert them into a matrix.

```{r}
expr_mat <- expr_wide |>
  select(-sample, -alive) |>
  as.matrix()

dim(expr_mat)
```

## Hierarchical clustering of samples

We calculate Euclidean distances between samples and perform hierarchical clustering with complete linkage.

```{r}
sample_dist <- dist(expr_mat)

hc_samples  <- hclust(sample_dist, method = "complete")

plot(
  hc_samples,
  main   = "Hierarchical clustering of samples",
  xlab   = "",
  sub    = "",
  labels = FALSE,
  hang   = -1
)


```

The hierarchical clustering reveals several transcriptomic subgroups across the TCGA-OV samples. However, these branches do not correspond clearly to the alive versus dead classification, confirming that global expression profiles are not primarily driven by vital status.

## Heatmap of top 100 variable genes

We reorder the matrix according to the clustering and generate a heatmap using base R. We also create a small annotation bar showing alive status for each sample.

```{r}
ordered_mat <- expr_mat[hc_samples$order, ]

annotation_alive <- expr_wide$alive[hc_samples$order]

row_colors <- ifelse(annotation_alive, "turquoise3", "salmon")

heatmap(
  ordered_mat,
  Rowv          = NA,
  Colv          = NA,
  scale         = "row",
  col           = colorRampPalette(c("navy", "white", "firebrick3"))(50),
  labRow        = FALSE,
  labCol        = FALSE,
  RowSideColors = row_colors,
  main          = "Heatmap of top 100 most variable genes"
)

```

The heatmap of the 100 most variable genes highlights strong co-expression patterns and substantial transcriptomic heterogeneity among patients. Despite this structure, the annotation bar shows no clear separation between alive and dead patients, suggesting that tumour biology drives the clustering more strongly than the clinical outcome.

## Conclusion

Although hierarchical clustering and heatmap analysis reveal clear transcriptomic subgroups within the TCGA-OV cohort, these groups do not align with the clinical variable of interest (alive vs dead).

The dominant structure reflects biological heterogeneity unrelated to survival, likely driven by differences in tumour microenvironment, batch effects, or molecular subtypes. Since this analysis does not provide additional insights into the clinical outcome we focus on in this project, we chose not to include these figures in the final presentation. Nevertheless, the clustering results are informative for understanding the global variability of the dataset and are retained in the project documentation for completeness.
